---
title: "Example of Supercell pipeline"
csl: elsevier-harvard.csl
output:
  github_document:
  bookdown::pdf_document2:
    number_sections: yes
abstract: |
  This is the abstract.
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.path = "figures/"
)
```

# Installation

```{r}
#devtools::install_github("mariiabilous/SCimple")
library(SCimple)

```

# Analysis
## Load data of 5 cancer cell lines sequenced with 10x from [Tian et al., 2019](https://doi.org/10.1038/s41592-019-0425-8)

```{r}
GE <- readRDS("./data/5cancer_cell_lines_10x_GE.Rds") # load gene expression matrix
dim(GE) # genes as rows and cells as columns
cell.meta <- readRDS("./data/5cancer_cell_lines_10x_cell_line_assignment.Rds") # load cell assignment to a cancer cell line
```
## Simplify single-cell data reducing numer of super-cell $gamma = 20$ times comparing to the number of single cells 
```{r Simplification, warning=FALSE, paged.print=FALSE}
gamma <- 20
SC    <- SCimplify(GE, k.knn = 5, gamma = gamma, n.var.genes = 1000, use.nn2 = FALSE)
set.seed(12345)
supercell_plot(SC$graph.supercells, main = paste("Super-cell network, gamma =", gamma), seed = 1)
supercell_plot(SC$graph.singlecell, group = cell.meta, do.frames = T, main = paste("Single-cell network, N =", dim(GE)[2]), lay.method = "components")
```

## Compute gene expression for simplified data
```{r}
SC.GE <- supercell_GE(GE, SC$membership)
dim(SC.GE)
```

#Map each super-cell to a particular cell line
```{r}
SC2cellline  <- supercell_assign(cell.meta, SC$membership)
SC$cell_line <- SC2cellline

supercell_plot(SC$graph.supercells, group = SC$cell_line, seed = 1, main = "Super-cell colored by cell line assignment")
```

Cluster super-cell data
```{r}
SC.PCA         <- supercell_prcomp(t(SC.GE), genes.use = SC$genes.use, supercell_size = SC$supercell_size, k = 20)
D              <- dist(SC.PCA$x)

SC.clusters    <- supercell_cluster(D = D, k = 5, supercell_size = SC$supercell_size)
SC$clustering  <- SC.clusters$clustering
```

```{r}
map.cluster.to.cell.line    <- supercell_assign(supercell_membership = SC$clustering, clusters  = SC$cell_line)
SC$clustering_reordered     <- map.cluster.to.cell.line[SC$clustering]
supercell_plot(SC$graph.supercells, group = SC$clustering_reordered, seed = 1, main = "Super-cell colored by cell line cluster")
```


Differential expression analysis of clustered super-cell data 
```{r}
markers.all.positive <- supercell_FindAllMarkers(ge = SC.GE, 
                                                 supercell_size = SC$supercell_size,
                                                 clusters = SC$clustering_reordered,
                                                 logfc.threshold = 1,
                                                 only.pos = T)
markers.all.positive$H1975[1:20,]
```

