---
title: "SuperCell 2.0 Tutorial: PBMC CITE-seq metacell atlas"
author: "Léonard Héraut"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    theme: readable
    fig_retina: 2
    self_contained: false
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10, fig.height = 6,
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
```

In this tutorial we will construct a multimodal metacell atlas of PBMC CITE-seq data using a semi-supervised workflow composed of SuperCell2.0 and STACAS.

The single-cell atlas was originally constructed and annotated in [Hao et al., 2021](https://doi.org/10.1016/j.cell.2021.04.048).

# Tutorial setup

Installing SuperCell2.0 package from github if needed.

```{r library, warning=FALSE, eval=FALSE}
if (!requireNamespace("remotes")) install.packages("remotes")
remotes::install_github("GfellerLab/SuperCell@supercell-2.0")
```

We will use STACAS for batch correction of the RNA and ADT modalities.

```{r}
if (!requireNamespace("STACAS")) {
  remotes::install_github("carmonalab/STACAS")
}

```

## Load libraries

```{r load library, warning=FALSE, message=FALSE}
library(SuperCell)
library(Seurat)
library(dplyr)
library(ggplot2)
library(patchwork)
library(STACAS)

```

## Download data from GEO

The two following chunks provide example command lines to download the atlas count matrices as well as atlas single-cell metadata from GEO (GSE164378) and place them locally in `"./PBMC_CITE_seq_atlas/input/GSM5008737_RNA_3P/"` and `"PBMC_CITE_seq_atlas/input/GSM5008738_ADT_3P/"`. The tutorial will use this path structure.

First, create input directory.

```{bash, eval = FALSE}
mkdir -p PBMC_CITE_seq_atlas/input &&
cd PBMC_CITE_seq_atlas/input &&
```

Then download data.

```{bash download_data, eval = FALSE}
cd PBMC_CITE_seq_atlas/input &&
wget "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE164378&format=file&file=GSE164378%5Fsc%2Emeta%2Edata%5F3P%2Ecsv%2Egz" -O GSE164378_sc.meta.data_3P.csv.gz &&
gunzip GSE164378_sc.meta.data_3P.csv.gz &&
wget "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE164378&format=file" -O GSE164378_RAW.tar &&
mkdir -p GSE164378_RAW && tar -C GSE164378_RAW -xvf GSE164378_RAW.tar &&
cd GSE164378_RAW &&
mkdir -p GSM5008737_RNA_3P && mv GSM5008737_RNA_3P-* GSM5008737_RNA_3P/ &&
mkdir -p GSM5008738_ADT_3P && mv GSM5008738_ADT_3P-* GSM5008738_ADT_3P/ &&
cd GSM5008737_RNA_3P/ &&
for file in GSM5008737_RNA_3P-*; do
mv "$file" "${file#GSM5008737_RNA_3P-}"
done &&
cd ../GSM5008738_ADT_3P/ &&
for file in GSM5008738_ADT_3P-*; do
mv "$file" "${file#GSM5008738_ADT_3P-}"
done 
```

In the end, some cleaning.

```{bash, eval = FALSE}
cd PBMC_CITE_seq_atlas/input &&
rm -f GSE164378_RAW/*.gz &&
mv GSE164378_RAW/* . &&
rm -r GSE164378_RAW
```

## Settings

```{r}
adt.comp <- 1:50
rna.comp <- 1:40
graining.level = 20
```

## Load single-cell data as a seurat object

This atlas contain more than 160,000 cells profiled with CITE-seq (RNA + surface proteins) using more than 200 antibody derived tags (ADTs).

```{r}
rna.counts <- Read10X("PBMC_CITE_seq_atlas/input/GSM5008737_RNA_3P/")
meta <- read.csv("PBMC_CITE_seq_atlas/input/GSE164378_sc.meta.data_3P.csv",row.names = 1)
pbmc <- CreateSeuratObject(counts = rna.counts,meta.data = meta)
remove(rna.counts)
remove(meta)
gc()

adt.counts <-  Read10X("PBMC_CITE_seq_atlas/input/GSM5008738_ADT_3P/")
pbmc[["ADT"]] <- CreateAssayObject(counts = adt.counts)
remove(adt.counts)
gc()

pbmc
```

This seurat single-cell object is annotated on different levels.

```{r}
head(FetchData(pbmc,c("donor","time","celltype.l1","celltype.l2")))
```

This atlas is composed of 8 donors profiled at 3 time points during the vaccine trial for a total of 24 samples.

```{r}
table(pbmc$donor,pbmc$time)
```

We can add the `"sample"` metadata by concatenating donor and time.

```{r}
pbmc$sample <- paste0(pbmc$donor,"_",pbmc$time)
```

We also add a prefix to the `"time"` metadata (in order to enable its aggregation at the metacell level with the `SCimplify_for_Seurat` function).

```{r}
pbmc$time <- paste0("time_",pbmc$time)
```

We can remove annotated doublets from the atlas.

```{r}
pbmc <- pbmc[,pbmc$celltype.l2 != "Doublet"]
# pbmc <- pbmc[,pbmc$celltype.l2 != "Doublet" & pbmc$donor %in% c("P1","P2","P3")]
pbmc
```

As in SuperCell2.0 manuscript, we add an intermediate level of annotation that we will used for the semi-supervised metacell workflow.

```{r}
pbmc$celltype.l1.5 <- pbmc$celltype.l1
pbmc$celltype.l1.5[pbmc$celltype.l1 == "other"] <- pbmc$celltype.l2[pbmc$celltype.l1 == "other"]
pbmc$celltype.l1.5[pbmc$celltype.l1 =="other T"] <- pbmc$celltype.l2[pbmc$celltype.l1 == "other T"]
```

# Metacell construction

We will process each single-cell sample separately for identifying metacells.

```{r}
pbmc.smp.list <- SplitObject(pbmc,split.by = "sample")
remove(pbmc)
gc()
```

We preprocess the single-cell modalities using standard Seurat workflows. Then metacells are identified using the semi-supervised multimodal SuperCell2.0 workflow with `"celltype.l1.5"` as input cell labels. This step can take around 15 min but could be further accelerated using parallelisation.

```{r message=FALSE}
options(future.globals.maxSize = 1.5 * 1024 ^ 3) 

t0 <- Sys.time()
pbmc.smp.list <- lapply(pbmc.smp.list, function(x) {
  #single-cell RNA preprocessing
  DefaultAssay(x) <- "RNA"
  x <- NormalizeData(x,verbose = FALSE) %>%
    FindVariableFeatures(verbose = FALSE) %>%
    ScaleData(verbose = FALSE)%>%
    RunPCA(verbose = FALSE)
  
  #single-cell ADT preprocessing
  DefaultAssay(x) <- "ADT"
  VariableFeatures(x) <- rownames(x)
  x <- NormalizeData(x, normalization.method = "CLR", margin = 2,verbose = FALSE) %>%
    ScaleData(verbose = FALSE) %>%
    RunPCA(reduction.name = "apca",verbose = FALSE)
  
  #metacell identificaiton with the semi-supervised multimodal SuperCell2.0 workflow
  x <- SCimplify_for_Seurat(
    seurat    = x,
    assay     = c("RNA", "ADT"),
    reduction = list("pca", "apca"),
    dims      = list(rna.comp, adt.comp),
    label     = "celltype.l1.5",
    gamma     = graining.level
  )
  return(x)
})

tf <- Sys.time()

tf - t0
```

`pbmc.smp.list` now contains seurat metacell objects that we can integrate using STACAS.

# Metacell integration

Here we will correct batch effect in each modality before performing a multimodal integration at the metacell level.

We define the samples at day 0 of the vaccine trial as the reference for integration of RNA and ADT modalities.

```{r}
reference <- which(endsWith(names(pbmc.smp.list),suffix = "_0"))
print(reference)
```

## Integration of the RNA modality

First we preprocess the metacell RNA data. We rename the metacells by adding a sample name prefix to avoid duplicated names at the integration step.

Accordingly, we also have to rename the metacell membership vector of each metacell object.

```{r message=FALSE}
pbmc.smp.list <- lapply(X = pbmc.smp.list, FUN = function(x) { 
  
  x <- RenameCells(x,add.cell.id = unique(x$sample))
  x@misc$new.membership <- paste0(unique(x$sample),"_Metacell_", x@misc$membership)
  names(x@misc$new.membership) <- names(x@misc$membership)
  DefaultAssay(x) <- "RNA"
  x <- NormalizeData(x, verbose = FALSE) %>% FindVariableFeatures()
  
  return(x)}
)
```

Then we integrate the RNA modality at the metacell level using the batch correction tool STACAS.

```{r message=FALSE}

features <- SelectIntegrationFeatures(object.list = pbmc.smp.list)

pbmc.combined <- Run.STACAS(object.list = pbmc.smp.list, 
                            assay = rep("RNA",length(pbmc.smp.list)),
                            anchor.features = features, 
                            dims = 50,  
                            cell.labels = "celltype.l1.5",
                            reference = reference)

stacas_rna  <- pbmc.combined[["pca"]]

```

## Integration of the ADT modality

First preprocessing of metacell ADT data.

```{r}

pbmc.smp.list <- lapply(X = pbmc.smp.list, FUN = function(x) {
  DefaultAssay(x) <- "ADT"
  VariableFeatures(x) <- rownames(x[["ADT"]])
  x <- NormalizeData(x, normalization.method = 'CLR', margin = 2) 
  
  return(x)
})
```

Then integration with STACAS.

```{r}

features <- rownames(pbmc.combined[["ADT"]]) # we use all ADTs

pbmc.combined <- Run.STACAS(object.list = pbmc.smp.list, 
                            assay = rep("ADT",length(pbmc.smp.list)),
                            anchor.features = features,
                            scale.data = T,
                            dims = 50,
                            cell.labels = "celltype.l1.5",
                            reference = reference)

pbmc.combined[["stacas_adt"]] <- pbmc.combined[["pca"]]
pbmc.combined[["pca"]] <- NULL
pbmc.combined[["stacas_rna"]] <- stacas_rna 

```

Finally we add a membership vector to the combined object.

```{r}
pbmc.combined@misc$membership <- unlist(lapply(pbmc.smp.list,FUN = function(x) {x@misc$new.membership}))
```


## Final metacell multimodal integration

```{r  fig.height=5,fig.width=8}

pbmc.combined <- FindMultiModalNeighbors(pbmc.combined,
                                         reduction.list = list("stacas_rna","stacas_adt"),
                                         dims.list = list(rna.comp,adt.comp),prune.SNN = 0,
                                         verbose = F)

pbmc.combined <- RunUMAP(pbmc.combined, 
                         nn.name = "weighted.nn", 
                         reduction.name = "wnn.umap", 
                         reduction.key = "wnnUMAP_",return.model = T) 

DimPlot(pbmc.combined,reduction = "wnn.umap",label = T,repel = T,group.by = c("celltype.l2"),shuffle = T)
DimPlot(pbmc.combined,reduction = "wnn.umap",group.by = c("sample"),shuffle = T)

```

This final embedding is slightly different from the one in our manuscript were we used SCTransform workflow to preprocess single-cell RNA modality (as in the original study) while here we use standard log normalization for both single-cell and metacell RNA processing in order to speed up the analysis. Overall the results are consistent.

## Comparison without batch correction

Without batch correction a batch effect is clearly present at the metacell level.

```{r fig.height=5,fig.width=8}
DefaultAssay(pbmc.combined) <- "RNA"
pbmc.combined <- ScaleData(pbmc.combined) %>% RunPCA()

DefaultAssay(pbmc.combined) <- "ADT"
VariableFeatures(pbmc.combined) <- rownames(pbmc.combined[["ADT"]])
pbmc.combined <- ScaleData(pbmc.combined) %>% RunPCA(reduction.name = "apca")

pbmc.combined <- FindMultiModalNeighbors(pbmc.combined, 
                        weighted.nn.name = "unintegrated.weighted.nn",
                        knn.graph.name = "unintegrated.wknn",
                        snn.graph.name = "unintegrated.wsnn",
                        reduction.list = list("pca","apca"),
                        dims.list = list(rna.comp,adt.comp),
                        verbose = F)

pbmc.combined <- RunUMAP(pbmc.combined, 
                         nn.name = "unintegrated.weighted.nn", 
                         reduction.name = "unintegrated.wnn.umap", 
                         reduction.key = "unintegrated.wnnUMAP_"
                         )

DimPlot(pbmc.combined,reduction = "unintegrated.wnn.umap",label = T,group.by = c("celltype.l2"),shuffle = T)
DimPlot(pbmc.combined,reduction = "unintegrated.wnn.umap",group.by = c("sample"),shuffle = T)
```

# Identification of interferon-primed CD14 monocytes

Let's perform an unsupervised clustering of our metacells based on the (batch corrected) multimodal metacell network we just obtained.

```{r}
DefaultAssay(pbmc.combined) <- "RNA"
pbmc.combined <- FindClusters(pbmc.combined,graph.name = "wsnn",resolution = 0.5,algorithm = 3)

DimPlot(pbmc.combined, reduction = "wnn.umap",label = T)
pbmc.combined$metacell_cluster <- pbmc.combined$wsnn_res.0.5
```

Cluster 0 corresponds to CD14 monocytes, cluster 2 to CD16 monocytes and cluster 16 is an unannotated cluster of CD14 monocytes expressing an interferon signature.

```{r fig.height=3,fig.width=6}
pbmc.combined[["RNA"]] <- JoinLayers(object = pbmc.combined[["RNA"]])
rna.vln <- VlnPlot.SuperCell(pbmc.combined[,pbmc.combined$metacell_cluster %in% c(0,2,16)],
                  features = c('CD14',"FCGR3A","ISG15","SIGLEC1","IFIT1"),
                  stack = T,fill.by = "ident") + xlab("RNA exp")

adt.vln <- VlnPlot.SuperCell(pbmc.combined[,pbmc.combined$metacell_cluster %in% c(0,2,16)],
                  features = c('CD14',"CD16"), assay = "ADT",
                  stack = T,fill.by = "ident") + xlab("Prot exp")

wrap_plots(adt.vln,rna.vln,guides = "collect",axis_titles = "collect",widths = c(2,4.5))
```

We can appreciate CD14 interferon-primed monocyte dynamic during the vaccine trial. These monocytes are present before the vaccination, their level decrease at during the trial before replenishing at the end.

```{r fig.height=6,fig.width=14}
DimPlot.SuperCell(pbmc.combined[,pbmc.combined$metacell_cluster %in% c(0,2,16) ],split.by = "time",reduction = "wnn.umap")
```
